The analysis of spatio-temporal data is an important part of research across many disciplines.
Spatio-temporal data is characterised by having both spatial, and temporal components that carry information across space, and over time.
Modelling these datasets requires thus observing and understanding how the phenomenon of interests evolves over time in many locations.
This subsection covers some basic principles of spatio-temporal data and dynamic modelling such data through the advection equation. 

\subsubsection{Defining Spatio-Temporal Data}
As to dismiss the Einstenian physics where space and time is interdependent, the classical way of separating the real spatial and temporal dimensions shall be considered here.

So now the spatial domain $\Omega$ is assumed to be a subset of $n$-dimensional real space, $\Omega \subset \R^n$, and the temporal domain $I$ a subset of the real line, $I \subset \R^1$.
Then the spatio-temporal domain is the Cartesian product of the spatial and temporal domains, $\Omega \times I$ and then an arbitrary spatio-temporal process can be written as
\begin{equation}
    \left\{ Y(x, t) \, \middle\vert \, x \in \Omega, \, t \in I \right\}.
\end{equation}

This way the spatio-temporal data may be thought as snapshots of some process at a specific point in time.
Naturally either, or both, the spatial and temporal variables may also be discrete by the nature of the quantities or due to discrete measurements.
In this thesis the developed model is tested with cloud optical thickness (COT) data.
The COT data used here consists of two-dimensional satellite images of the passing of light trough the atmosphere at geographical coordinates (latitude, longitude) in 15-minute time intervals.


\subsubsection{Dynamic Modelling of Spatio-Temporal Data}
Dynamical modelling of spatio-temporal data refers to creating mathematical and computational models that explain changes of the observed variables in time.
Often some prior scientific information of the process is used to motivate the model.
This information is usually described with some partial differential equations (PDE).
At the core of the model of this thesis is the advection equation describing the transportation of the observed variable along some advection field in time.
So in the COT example one could intuitively think that clouds move according to some winds that create an advection field.

Consider a spatio-temporal function $u: \Omega \times I \rightarrow \R$ representing some quantity on spatial domain $\Omega \subset \R^n$, and temporal domain $I = (0, T)$.
The advection equation is now defined as
\begin{equation}\label{eq:advection}
    \frac{\partial u}{\partial t} =  - \mathbf{F} \nabla u,
\end{equation}
where $\mathbf{F} = \mathbf{F}(x,t)$ is the vector-valued advection field that transport $u$ forward in time.

If the advection field $\mathbf{F}$ was known, and the model was perfect $u$ could be solved given some initial condition $u(\cdot, 0) = u_0(\cdot)$, and a boundary condition for the values of $u$ in the Dirichlet type case or the normal derivative of $u$ along the boundary of $\Omega$ in the Neumann type case.
However, measuring advection and diffusion parameters is often not possible and estimates of $\overline{\mathbf{F}}$ need to be extracted e.g.\ from data.
While many uncertainties are involved in the parameter estimation, the model~\eqref{eq:advection} is also not a perfect representation so the approximate stochastic model
\begin{equation}\label{eq:advection_approx}
    \frac{\partial u}{\partial t} = - \overline{\mathbf{F}} \nabla u + e
\end{equation}
needs to be considered where $e = e(x,t)$ is the (hopefully small) modeling error.
When taking real measurements, the model also needs to be discretised both in time so that each time step represents one measurement point, and in space so that $\Omega$ becomes a grid in which the measurements are taken at some resolution.

In the COT example the motivation for computing the estimates of $\overline{\mathbf{F}}$ would be that assuming the real advection field changes little in time, i.e.\ $\mathbf{F}(x,t_{k}) \approx \mathbf{F}(x,t_{k+1})$, it can be utilised to transport COT images $u$ forward in time. 
