While the described model is designed to be data-independent, it was primarily developed for the Cloud Optical Thickness (COT) data retreived from the European Organisation for the Exploitation of Meteorological Satellites' (EUMETSAT) geostationary satellites.
COT observations provide a consistent data source suitable for the model.
The choice of using COT data is motivated by two key factors; (1) it is easiliy available with regular capturing suitable for short-term forecasting applications, (2) the nature of the COT data makes it suitable for the studied advection modelling.

COT, as a measure of a cloud's ability to absorb and scatter light at visible wavelengths is an essential parameter in satellite observations of clouds.
Within the EUMETSAT data, it is a part of the Optimal Cloud Analysis (OCA) product, derived from optical Flexible Combined Imager instruments measuring 16 channels of visible and infrared spectrum of light~\cite{oca, fci}.
The OCA algorithm defines a state of a cloudy atmosphere as a state of six parameters of which COT is one that measures a cloud's ability to absorb and scatter mostly at visible wavelengths of light at daytime.
The COT data range is 0.063 to 256 limited by the lookup table used by the OCA algorithm.

The data is captured by EUMETSAT's geostationary satellites at 36000 km above the equator, which remain at a constant position relative to the Earth.
The stationarity enables continuous observations of the same geographical region, which is indispensable for retrieving coherent spatio-temporal data.
However, due to the geostationary orbit above the equator, the resolution of the data varies and the quality of the data is poorer at higher latitudes.
This dataset has a temporal resolution of 15 minutes, enabling accurate tracking of the evolution of clouds.

For a comprehensive understanding of the algorithmic principles behind retrieval of the COT data, readers are referred to~\cite{oca}.
